---
title: "Epigenetic age in ML10"
author: "ks"
date: "2023-10-31"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r}
#devtools::install_github("MorganLevineLab/prcPhenoAge")
#devtools::install_github("danbelsky/DunedinPoAm38")
#devtools::install_github("MorganLevineLab/methylCIPHER")
if (!require("sesame")) { BiocManager::install("sesame"); library("sesame")}
```


```{r libraries, message=FALSE}
library(readxl)
library(dplyr)
library(purrr)
library(sesame)
library(ggplot2)
library(tidyr)
library(methylCIPHER)
library(patchwork)
```

# Pilot Data: ML3, ML10, MCV39  

Let's read in the MethylationEPICV2 data.

```{r read-dat, cache=TRUE}
#idat_dir = c("~kims/Google Drive/My Drive/Data/Dubeau/1756_Dubeau EPICv2 DNA methylation data package/IDAT FILES")
idat_dir = c("/Volumes/extreme ssd/Data/Dubeau/1756_Dubeau EPICv2 DNA methylation data package/IDAT FILES")
betas = openSesame(idat_dir)
dim(betas)
save(betas,file="data/betas.rda")
```

How many different probe types?
```{r}
table(substr(rownames(betas),1,2))
```

```{r missingbetas}
apply(is.na(betas),2,mean)
```

```{r}
rownames(betas) <- substr(rownames(betas),1,10)
```



```{r read-phenotypes}
pdata = read_excel(c("/Volumes/extreme ssd/Data/Dubeau/1756_Dubeau EPICv2 DNA methylation data package/SAMPLE-ARRAY MAPPING/1756 (Dubeau-8).xlsx"))
sampleinfo <- strsplit(pdata$`Sample ID`, ', p')
AgeData <- cbind.data.frame(Barcode = pdata$`Complete Barcode`,
                      cell.line = map_chr(sampleinfo,pluck,1),
                      passage = as.numeric(map_chr(sampleinfo,pluck,2)),
                      gp=rep(1:4,each=2))
AgeData
```

## EpiTOC 

How many clock loci are missing?
```{r}
getClockProbes(t(betas[,3]))
```

We can investigate the missing, but with only 5 features out of 163 affected for EpiToc2, I'll ignore it for now.

Compute EPITOC2 and EPITOC:

```{r compute-epitoc2}
epitoc2 <- calcEpiTOC2(t(betas), AgeData, imputation = F)
```

```{r compute-epitoc}
epitoc <- calcEpiTOC(t(betas), epitoc2 , imputation = F)
epitoc
```

Now plot EPITOC2 for ML10.

```{r display}
dfx <- epitoc %>%
         filter(cell.line=="ML10")
#dfx
save(dfx, file="data/epitoc.rda")
```

Plot EPITOC2 for ML10.

```{r plot}
p1 <- ggplot(dfx,  aes(x=passage, y=epiTOC2, color = factor(gp))) +
          geom_point(size=2) +
          geom_line(lwd=1)  + ggtitle("epiTOC2") + 
          theme(legend.position="none")
p1
```

Is the pattern the same for EPITOC?

```{r display-epitoc}
p2 <- ggplot(dfx,  aes(x=passage, y=EpiTOC, color = factor(gp))) +
          geom_point(size=2) +
          geom_line(lwd=1) + ggtitle("EpiTOC") + theme(legend.position="none")
p2
```

Now let's add stemTOC
```{r}
#devtools::install_github("aet21/EpiMitClocks")
library("EpiMitClocks")
```


# this computes it from methylCIPHER package:
```{r compute-epitoc2}
epitoc2 <- calcEpiTOC2(t(betas), AgeData, imputation = F)
```


```{r}
EpiMitClocks.o<-EpiMitClocks(data.m = betas)
```

```{r}
EpiMitClocks.df <- data.frame(epiTOC1 = EpiMitClocks.o$epiTOC1,
                              epiTOC2 = EpiMitClocks.o$epiTOC2$tnsc,
                         stemTOCvitro = EpiMitClocks.o$stemTOCvitro,
                         stemTOC      = EpiMitClocks.o$stemTOC)
EpiMitClocks.df
```
```{r}
identical(epitoc$epiTOC2, EpiMitClocks.df$epiTOC2)
identical(epitoc$EpiTOC, EpiMitClocks.df$epiTOC1)
```

Phew. The two packages replicate, as they should. yay!

Now combine and plot the new stemTOC clocks for ML10
```{r}
allclocks <- cbind.data.frame(epitoc,
                              stemTOCvitro = EpiMitClocks.df$stemTOCvitro,
                              stemTOC      = EpiMitClocks.df$stemTOC)
allclocks
```


```{r display}
dfx <- allclocks %>%
         filter(cell.line=="ML10")
#dfx
save(dfx, file="data/epitoc.rda")
```


Show the growth changes using stemTOCs
```{r display-stemTOCs}
p3 <- ggplot(dfx,  aes(x=passage, y=stemTOC, color = factor(gp))) +
          geom_point(size=2) +
          geom_line(lwd=1)  + ggtitle("stemTOC") + 
          theme(legend.position="none")
p3
```


```{r display-stemTOCs}
p4 <- ggplot(dfx,  aes(x=passage, y=stemTOCvitro, color = factor(gp))) +
          geom_point(size=2) +
          geom_line(lwd=1)  + ggtitle("stemTOCvitro") + 
          theme(legend.position="none")
p4
```

```{r}
(p1+p2)/(p3+p4)
```


```{r sI}
sessionInfo()
```

